The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-03-10 paper/forestecology.Rmd
According to the koRpus method this is at 2994! Feel free to reject these changes if you don't like them. Or happy to chat about them tomorrow.

abstract: 
- |
  | 1. Neighborhood competition models are powerful tools to measure the effect of interspecific competition. Statistical methods to ease the application of these models are currently lacking.
  | 2. We present the `forestecology` package providing methods to i) specify neighborhood competition models, ii) evaluate the effect of competitor species identity using permutation tests, and iii) measure model performance using spatial cross-validation.  Following @allen_permutation_2020, we implement a Bayesian linear regression neighborhood competition model.
  | 3. We demonstrate the package's functionality using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO global network of research sites. Given ForestGEO's data collection protocols and data formatting standards, the package was designed with cross-site compatibility in mind. We highlight the importance of spatial cross-validation when interpreting model results. 
  | 4. The package features i) `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence, ii) functions with standardized inputs/outputs of simple features `sf` package class, and iii) an S3 object-oriented implementation of the Bayesian linear regression model. These three facts allow for clear articulation of all the steps in the sequence of analysis and easy wrangling and visualization of the geospatial forestry data. Furthermore, while the package only has Bayesian linear regression implemented, the package was designed with extensibility to other methods in mind.

@@

Repeat-censused forest plots offer excellent opportunities to test neighborhood models of the effect of competition on the growth of trees (@canham_neighborhood_2004). Neighborhood models of competition have been used to: test whether the species identity of a competitor matters (@uriarte_spatially_2004); measure species-specific competition coefficients (@das_effect_2012 @tatsumi_estimating_2016); test competing models to see what structures competitive interactions, e.g. traits or phylogeny (@allen_permutation_2020; @uriarte_trait_2010); and inform selective logging practices (@canham_neighborhood_2006). Although these are well-described methods, few methods are currently available for easy application.

We address this shortcoming with the `forestecology` R package providing methods and data for forest ecology model fitting and assessment, available on CRAN (https://cran.r-project.org/package=forestecology) and on GitHub (https://github.com/rudeboybert/forestecology). The package is written to model stem radial growth between two censuses based on neighborhood competition, largely following the methods in @allen_permutation_2020.

Let $i = 1, \ldots, n_j$ index all $n_j$ trees of "focal" species group $j$; let $j = 1, \ldots, J$ index all $J$ focal species groups; and let $k = 1, \ldots, K$ index all $K$ "competitor" species groups. The average annual growth in diameter at breast height (DBH) $y_{ij}$ (in centimeters/year) of the $i^{th}$ tree of focal species group $j$ is modeled as

@@

where $\beta_{0,j}$ is the diameter-independent growth rate for group $j$; $\text{dbh}_{ij}$ is the DBH of the focal tree at the earlier census for group $j$ and $\beta_{\text{dbh},j}$ quantifies its corresponding growth rate change;  $\text{BA}_{ijk}$ is the sum of the basal area of all trees of competitor species group $k$ and $\lambda_{jk}$ quantifies the corresponding change in growth for individuals of group $j$ from these competitors; and $\epsilon_{ij}$ is a random error term distributed $\text{Normal}(0, \sigma^2)$. They estimate all parameters via Bayesian linear regression, while exploiting Normal/Inverse Gamma conjugacy to derive closed-form solutions to all posterior distributions^[See S1 Appendix of @allen_permutation_2020, available at https://doi.org/10.1371/journal.pone.0229930.s004]. These closed-form solutions are not as computationally expensive as approximations from Markov Chain Monte Carlo algorithms.

@@

We present a case-study of `forestecology`'s functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot in Front Royal, VA, USA, part of the ForestGEO global network of research sites [@bourg_initial_2013, @andersonteixeira_ctfs-forestgeo_2015] [@bourg_initial_2013]. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US---the Blue Ridge, Ridge and Valley, and Piedmont provinces---and is adjacent to the northern end of Shenandoah National Park.

@@

```{r scbi-trees, out.width="66%", fig.cap="Compute growth of trees based on census data. A map of the growth of a random sample of 500 trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot."}
ggplot() +
  geom_sf(data = growth_scbi %>% sample_n(500), aes(size = growth)) + 
  scale_size_binned(limits = c(0.1, 1))
```

@@

We observe 4 variables describing 20 competitor trees: the unique tree-stem ID, the distance to the focal tree (all $\leq$ `r comp_dist`), the species, and the basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ for the DBH in cm from the earlier census. Saving competitor information in list-column format minimizes redundancy since we do not need to repeat information on the focal tree 20 times. We visualize the spatial distribution of these trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

@@

For many users the visualizations of $\lambda_{j,k}$ will be of particular interest as they provide insight into species competitive interactions, where negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip poplars (litu) have a strong negative effect on the growth of conspecifics but relatively lesser effect on pignut hickory (cagl) and red oak (quru) neighbors.

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model performance for 49 permutations of competitor species and RMSE calculations, both with and without cross-validation. Without cross-validation, competitor species identity does matter as the observed RMSE was significantly lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot, competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020 @uriarte_spatially_2004]. Furthermore, the larger cross-validated RMSE's are indicative of the importance of being vigilant against potential model overfitting.

@@

The `forestecology` package provides an accessible way to fit and test models of neighborhood competition. While the package is designed with ForestGEO plot data in mind, we envision that it can be modified to work on any single large, mapped forest plot in which at least two measurements of each individual have been taken, e.g. the US Forest Service Forest Inventory and Analysis plots. Furthermore, we also envision the package being used more generally to model interactions of any community of mapped sessile organisms [@smith_forest_2002]. In future versions of `forestecology` we also hope to include models that account for tree mortality in addition to tree growth. The package follows the `tidy` data design principles, leverages the `sf` package for spatial data, and S3 open-oriented model implementation structure. We hope that the package will increase the use of neighborhood competition models to better understand what structures plant competition.

>>>>

abstract: 
- |
  | 1. Neighborhood competition models are powerful tools to measure the effect of interspecific competition. Statistical methods to ease the application of these models are currently lacking.
  | 2. We present the `forestecology` package providing methods to i) specify neighborhood competition models, ii) evaluate the effect of competitor species identity using permutation tests, and iii) measure model performance using spatial cross-validation.  Following @allen_permutation_2020, we implement a Bayesian linear regression neighborhood competition model.
  | 3. We demonstrate the package's functionality using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO global network of research sites. Given ForestGEO's data collection protocols and data formatting standards, the package was designed with cross-site compatibility in mind. We highlight the importance of spatial cross-validation when interpreting model results. 
  | 4. The package features i) `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence, ii) functions with standardized inputs/outputs of simple features `sf` package class, and iii) an S3 object-oriented implementation of the Bayesian linear regression model. These three facts allow for clear articulation of all the steps in the sequence of analysis and easy wrangling and visualization of the geospatial data. Furthermore, while the package only has Bayesian linear regression implemented, the package was designed with extensibility to other methods in mind.

@@

Repeat-censused forest plots offer excellent opportunities to test neighborhood models of the effect of competition on the growth of trees [@canham_neighborhood_2004]. Neighborhood models of competition have been used to: test whether the species identity of a competitor matters [@uriarte_spatially_2004; measure species-specific competition coefficients [@das_effect_2012 @tatsumi_estimating_2016]; test competing models to see what structures competitive interactions, e.g. traits or phylogeny [@allen_permutation_2020; @uriarte_trait_2010]; and inform selective logging practices [@canham_neighborhood_2006]. Although these are well-described methods, few methods are currently available for easy application.

We address this shortcoming with the `forestecology` R package providing methods and data for forest ecology model fitting and assessment, available on CRAN (https://cran.r-project.org/package=forestecology) and on GitHub (https://github.com/rudeboybert/forestecology). The package is written to model stem diameter growth between two censuses based on neighborhood competition, largely following the methods in @allen_permutation_2020.

Let $i = 1, \ldots, n_j$ index all $n_j$ trees of "focal" species $j$; let $j = 1, \ldots, J$ index all $J$ focal species; and let $k = 1, \ldots, K$ index all $K$ "competitor" species. The average annual growth in diameter at breast height (DBH) $y_{ij}$ (in centimeters/year) of the $i^{th}$ tree of focal species $j$ is modeled as

@@

where $\beta_{0,j}$ is the diameter-independent growth rate of species $j$; $\text{dbh}_{ij}$ is the DBH of the focal tree at the earlier census and $\beta_{\text{dbh},j}$ the slope of that species's diameter-growth relationship;  $\text{BA}_{ijk}$ is the sum of the basal area of all trees of competitor species $k$ and $\lambda_{jk}$ quantifies the corresponding change in growth for individuals of group $j$ from these competitors; and $\epsilon_{ij}$ is a random error term distributed $\text{Normal}(0, \sigma^2)$. @allen_permutation_2020 estimate all parameters via Bayesian linear regression, while exploiting Normal/Inverse Gamma conjugacy to derive closed-form solutions to all posterior distributions^[See S1 Appendix of @allen_permutation_2020, available at https://doi.org/10.1371/journal.pone.0229930.s004]. These closed-form solutions are not as computationally expensive as approximations from Markov Chain Monte Carlo algorithms.

@@

We present a case-study of `forestecology`'s functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot in Front Royal, VA, USA, part of the ForestGEO global network of research sites [@bourg_initial_2013, @andersonteixeira_ctfs-forestgeo_2015]. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US---the Blue Ridge, Ridge and Valley, and Piedmont provinces---and is adjacent to the northern end of Shenandoah National Park.

@@

```{r scbi-trees, out.width="66%", fig.cap="Compute growth of trees based on census data. A map of the growth of a random sample of 500 trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot."}
ggplot() +
  geom_sf(data = growth_scbi %>% sample_n(500), aes(size = growth)) + 
  scale_size_binned(limits = c(0.1, 1)) +
  labs(size = expression(paste(Growth, " (cm ",y^{-1},')')) )
```

@@

We observe 4 variables describing 20 competitor trees: the unique tree-stem ID, the distance to the focal tree (all $\leq$ `r comp_dist` m), the species, and the basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ for the DBH in cm from the earlier census. Saving competitor information in list-column format minimizes redundancy since we do not need to repeat information on the focal tree 20 times. We visualize the spatial distribution of these trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

@@

For many users the visualizations of $\lambda_{j,k}$ will be of particular interest as they provide insight into species-specific competitive interactions, where negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip poplars (litu) have a strong negative effect on the growth of conspecifics but relatively lesser effect on pignut hickory (cagl) and red oak (quru) neighbors.

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model performance for 49 permutations of competitor species and RMSE calculations, both with and without cross-validation. Without cross-validation, competitor species identity does matter as the observed RMSE was significantly lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot, competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020 @uriarte_spatially_2004]. This provides a striking example of the importance of cross-validation, as without it the over-fit model gives rise to an incorrect conclusion.

@@

The `forestecology` package provides an accessible way to fit and test models of neighborhood competition. While the package is designed with ForestGEO plot data in mind, we envision that it can be modified to work on any single large, mapped forest plot in which at least two measurements of each individual have been taken. Furthermore, we also envision the package being used more generally for inventory plot-structure with many spatially separated plots, e.g. the US Forest Service Forest Inventory and Analysis plots [@smith_forest_2002]. In future versions of `forestecology` we also hope to include models that account for tree mortality in addition to tree growth. The package follows the `tidy` data design principles, leverages the `sf` package for spatial data, and S3 open-oriented model implementation structure. We hope that the package will increase the use of neighborhood competition models to better understand what structures plant competition.
